---
title: "Heatmap Lipid Cytokines"
author: "Katie"
date: "September 12, 2019"
output:
  pdf_document: default
  html_document: default
---

### import packages
```{r, message=FALSE}
library(tidyverse)
library(pheatmap)
library(readr)
```

### read in data
```{r, message=FALSE}
# reading in cytokine data
cytokine18 <- read_csv("cytokines0619.csv", na = "N/A")

# reading in lipid data
lipid18 <- read_csv("lipids0619.csv", na = 'N/A')

#reading in clinical data
clinic18 <- read_csv("clinical819_rev.csv")
```

### merge lipids and cytokines
```{r}
alldata <- inner_join(cytokine18, lipid18, by = c("SampleID", "LGLLType",  "Age", "Gender", "Immuno", "STAT3mut", "STAT3status", "Statin", "FishOil"))
```

### Merge all data with clinical data
```{r}
alldata <- left_join(alldata, clinic18, by = c("SampleID",  "Gender"))
```

We decided to impute with the median for all of these variables because distributions can be skewed for LGL patients

```{r}
alldata %>%
  #filter(Immuno == "N") %>%
  group_by(LGLLType) %>%
  summarise(med_sICAM = median(sICAM_1, na.rm = TRUE), 
            med_sFASlig = median(sFas_Ligand, na.rm = TRUE), 
            med_sFAS = median(sFas, na.rm = TRUE), 
            med_sVCAM1 = median(sVCAM_1, na.rm = TRUE))
```

Now impute using the above medians for normal patients. 
These values do not change with / without Immuno line because patients taking immunosuppression drugs were leukemic patients rather than normals

```{r}
alldata$sICAM_1[71] <- 150931 #patient 73 is at row 71
alldata$sFas_Ligand[71]<- 75
alldata$sFas[71]<-6192
alldata$sVCAM_1[71]<- 777563
```

Check for other missing values on biomarkers

```{r}
alldata %>%
  dplyr::select(MIG:SMC26_1) %>%
  is.na() %>%
  sum()
```
No other missing values

###Remove Patients on immunosuppressive treatment

**ALSO remove patient ID == 20**
Patient 20 was excluded because they were initially incorrectly identified and were found to be on immunosuppressive treatment
```{r}
alldata_noImmuno <- alldata %>%
  filter(Immuno == "N") %>%
  filter(SampleID != 20)
rm(alldata)
```

# Pre-processing for Heatmap

### Create biomarker dataframe
selecting the biomarkers from alldata_noImmuno

```{r}
biomarkers_noImmuno <- alldata_noImmuno %>%
  dplyr::select(MIG:SMC26_1)

# drop tgfb3 because it has 0 variance (all observations below limit of detection)
biomarkers_noImmuno <- biomarkers_noImmuno %>%
  dplyr::select(-TGFB3)
```

### log10 transform of data
```{r}
biomarkers_noImmuno <- log10(biomarkers_noImmuno)
```

### scale data
```{r}
biomarkers_noImmuno <- scale(biomarkers_noImmuno)
```

### set row names for main matrix values
```{r}
rownames(biomarkers_noImmuno) <- alldata_noImmuno$SampleID

write.csv(biomarkers_noImmuno, file = "biomarkers_noImmuno.csv")
```

### make col/row annotation dataframes
Because we want to annotate the heatmap with these variables, we're making a dataframe of them.
**Ensure that the annotation dataframe and the biomarker dataframe have the same patient order**
```{r}

annrow_df <- alldata_noImmuno %>%
  select(Gender, LGLLType, STAT3status, MCV, Hgb, ANC)
rownames(annrow_df) <- alldata_noImmuno$SampleID
annrow_df$Gender <- factor(annrow_df$Gender)
annrow_df$LGLLType <- factor(annrow_df$LGLLType)
annrow_df$STAT3status <- factor(annrow_df$STAT3status)

write.csv(annrow_df, file = "annotations.csv")
```

#Create the Heatmap
```{r}
#loads in the files for plotting and set sampleID (currently column 1) as the rownames
biomarkers_noImmuno <- read.csv("biomarkers_noImmuno.csv", row.names = 1)
sample_ann_df <- read.csv("annotations.csv", row.names = 1)
```

```{r}
#sets annotation colors
ann_colors = list(Gender = c(M = 'turquoise', F = 'plum'), 
                  LGLLType = c("NK-LGLL" = "gold", "Normal Donor" = 'orchid1', "T-LGLL" = "deepskyblue"), 
                  STAT3status = c(D661I = 'deepskyblue', D661Y = 'palegreen', K658R = 'pink', N647I = 'darkorange', S614R = 'palegoldenrod', WT = 'azure2', Y640F = 'darkorchid', 'Y640F, D661Y' = 'royalblue', 'Y640F, I659L' = 'violet', 'Y640F, Q643H' = 'black'))

#plots heatmap
options('repr.plot.height' = 12, 'repr.plot.width' = 12)
p <- pheatmap(t(biomarkers_noImmuno),
              color = colorRampPalette(c("blue", "white", "red"))(n = 1000),
              clustering_distance_rows = "euclidean",
              clustering_distance_cols = 'euclidean',
              annotation_col = sample_ann_df,
              annotation_colors = ann_colors,
              main = 'Lipid and Cytokine Clustered Heatmap', 
              border_color = NA,
              show_colnames = FALSE,
              height = 9,
              width = 9.5,
              filename = "newheatmap.png")

```


